A numerical study of variational data assimilation for the wave equation in heterogeneous media

### 2024-05-14 tex/main.tex
Updates from Overleaf
Minor

\affil[2]{\orgdiv{Institute for Numerical and Applied Mathematics}, \orgname{University of G\"{o}ttingen}, \orgaddress{\street{Lotzenstr.}, \city{G\"{o}ttingen}, \postcode{37083}, \country{Germany}}}

@@

\section{Introduction}
\noindent In this article, we consider the discretization of a data assimilation problem for the wave equation in heterogeneous media. In the following, let $\Omega \subset \mathbb{R}^d$ be a bounded domain with smooth boundary $\partial \Omega$ such that there exists subdomains $\Omega_1, \Omega_2 \subseteq \Omega$ such that $\Omega = \overline{\Omega}_1 \cup \overline{\Omega}_2$ and $\Omega_1 \cap \Omega_2 = \emptyset$. We consider a scalar wave speed $c(x) = \mathbbm{1}_{\Omega_1} c_1(x) + \mathbbm{1}_{\Omega_2} c_2(x)$, $c_i \in C^\infty(\Omega_i)$, $i = 1,2$, which may be discontinuous across the interface $\Gamma = \Omega \setminus (\Omega_1 \cup \Omega_2)$, but constant in time. For a final time $T > 0$, we consider the space-time domain $Q = (0,T) \times \Omega$ with lateral boundary $\Sigma \coloneqq (0,T) \times \partial \Omega$. For any nonempty subset $\omega \subset \Omega$, we set $\omega_T \coloneqq (0,T) \times \omega$. In the following, let $\wop$ be the wave operator defined as 
\begin{equation*}
    \wop u := \partial_t^2 - \div(c^2 \nabla u).
\end{equation*}
In particular, we write $\square_1$ if $c_i = 1$, $i = 1,2$, i.e. if we are considering the case of homogeneous media. 
Then, we consider the problem: find $u : Q \rightarrow \mathbb{R}$ such that 
\begin{equation}\label{eq:waveEquation}
        (\wop u, u) = (0,0) \quad \text{ in } Q \times \Sigma.  
\end{equation}
To ensure that the problem is well posed, one usually prescribes initial data
\begin{equation}\label{eq:initialData}
    (u,\dt u) \vert_{t = 0} = (u_0,u_1) \text{ in } \Omega, \tag{IVP} 
\end{equation}
where $u_0 \in H^1(Q)$ and $u_1 \in L^2(Q)$ are given. For a detailed analysis of the well-posedness of the problem \eqref{eq:waveEquation}-\eqref{eq:initialData} in heterogeneous media we refer to \cite{StolkPhD}. 
Here we will consider a data assimilation problem instead, i.e. we assume that the initial data is unknown and that there are given measurements $u_{\omega} \in L^2(\omega_T)$ which are prescribed in the data domain $\omega_T$: 
\begin{equation}\label{eq:dataMatch}
    u = u_{\omega} \text{ in } \omega_T. \tag{DA}
\end{equation}

@@

\noindent For the heterogeneous case where $c_1 \not =  c_2$, the problem is less studied. In the absence of the GCC, we have to assume that the final time $T$ is large enough to ensure that problem \eqref{eq:waveEquation}-\eqref{eq:dataMatch} is well-posed. To be precise, Holmgren's unique continuation theorem guarantees that \eqref{eq:waveEquation}-\eqref{eq:dataMatch} is uniquely solvable if
\begin{equation}\label{eq:Tcondition}
    T > 2 \sup_{x \in \Omega} \operatorname{dist}_{c}(x,\omega_T),
\end{equation}
where $\operatorname{dist}_c(x,\omega_T) := \inf_{y \in \omega_T} d_c(x,y)$ with $d_c$ being the distance function defined as the length of the shortest continuous path in $\Omega$ connecting $x$ and $y$ with respect to the metric adapted to the discontinuous wave speed $c$. While the condition \eqref{eq:Tcondition} guarantees unique solvability, the stability of problem \eqref{eq:waveEquation}-\eqref{eq:dataMatch} can be poor. In fact, it has been shown \cite[Thm. 5.17]{Filippas22} that the assumption \eqref{eq:Tcondition} only guarantees the following result:  
\begin{thm}[Thm. 5.17 of \cite{Filippas22}]\label{thm:filippas}
    Let $\omega_T \subset \Omega$ be nonempty and let $T$ be s.t. \eqref{eq:Tcondition} holds. Then, there exists $C,\kappa$ such that for any initial data $(u_0,u_1) \in H^1_0(\Omega) \times L^2(\Omega)$ and $u$ solving \eqref{eq:waveEquation}-\eqref{eq:initialData} one has for any $\mu > 0$ that 
    \begin{equation}\label{eq:FilippasEstimate}
        \Vert (u_0,u_1) \Vert_{L^2 \times H^{-1}} \le C e^{\kappa \mu} \left( \Vert u \Vert_{L^2(\omega_T)} + \Vert \wop u \Vert_{L^2(Q)} \right) + \frac{C}{\mu} \Vert (u_0,u_1) \Vert_{H^1 \times L^2}. 
    \end{equation}
    In particular, if $(u_0,u_1) \not = (0,0)$ we have that 
    \begin{equation}\label{eq:FilippasEstimateLog}
        \Vert (u_0,u_1) \Vert_{L^2 \times H^{-1}} \le C \frac{\Vert (u_0,u_1) \Vert_{H^1 \times L^2}}{\log \left( 1 + \frac{\Vert (u_0,u_1) \Vert_{H^1 \times L^2}}{ \Vert u \Vert_{L^2(\omega_T)} + \Vert \wop u \Vert_{L^2(Q)}} \right)}. 
    \end{equation} 
\end{thm}

@@

\section{Discretization (fully discrete \& dG in time)} 
\noindent In this section, we introduce a fully discrete discontinuous in time finite element method to discretize the data assimilation problem \eqref{eq:waveEquation}-\eqref{eq:dataMatch} for the wave equation in heterogeneous media. The discretization is based on the scheme introduced in \cite{BP24} for the homogeneous case. Before introducing the method in Section \ref{sec:method}, we introduce a geometric partition of the space-time domain $Q$ into time slabs in Section \ref{sec:spaceTimeDiscretization}.

\subsection{Discretization of the space-time domain}\label{sec:spaceTimeDiscretization}
Let $\mathcal{T}_h$ be a quasi-uniform triangulation of $\Omega$ such that both, the data domain $\omega$ and the interface $\Gamma$ are fitted. 
Note that the triangulation can be constructed such that for $h$ small enough the following continuous trace inequality holds \cite[Sec. 4.2]{BFMO21control}: For all $v \in [H^1(K)]^d$ and $K \in \mathcal{T}_h$, we have
\begin{equation}\label{eq:traceInequality}
    \Vert v \Vert_{[L^2(\partial K)]^d} \le C \left(h^{-1/2} \Vert v \Vert_{[L^2(K)]^d} + h^{1/2} \Vert \nabla v \Vert_{[L^2(K)]^d} \right).  
\end{equation}

@@

\section{Error analysis}
\noindent In this section, we analyze the approximation of the continuous problem \eqref{eq:waveEquation}-\eqref{eq:dataMatch} through \eqref{eq:discreteProblem}. In particular, we show that the error between the first component $\ul_1$ of the discrete solution and the continuous solution $u$ is bounded by the best approximation error in suitable discrete norms defined below. Then, we show that under Assumption \ref{assum:LipschitzStability}, we can recover convergence rates in a physically relevant norm.
Let $\vert \cdot \vert_{S_h}, \vert \cdot \vert_{\uparrow \downarrow}$, and $\Vert \cdot \Vert_{S_h^{\ast}}$ be the semi-norms, respectively norms, induced by the stabilizers $S_h$, $\Sud$, and $S_h^{\ast}$:
\begin{equation}
    \vert \Uh \vert^2_{S_h} := S_h(\Uh,\Uh), \quad \vert \Uh \vert^2_{\uparrow \downarrow} := \Sud(\Uh,\Uh), \quad \Vert \Zh \Vert^2_{S_h^{\ast}} := S_h^{\ast}(\Zh,\Zh).
\end{equation}
\noindent Then, we define the discrete norm 
\begin{equation}
    \tnorm{ (\Uh,\Zh) }^2 := \gamma \vert \Uh \vert^2_{S_h} + \vert \Uh \vert^2_{\uparrow \downarrow} + \Vert \ul_1 \Vert^2_{\omega_T} + \gamma^{\ast} \Vert \Zh \Vert^2_{S_h^\ast},
\end{equation}
and its strengthened version
\begin{equation}\label{eq:tnormwop}
    \begin{aligned}
        \tnorm{(\Uh, \Zh)}^2_{\wop} := \tnorm{(\Uh, \Zh)}^2 &+ \sum_{n = 0}^{N -1} \Big\{ \Vert \dt u_2 \Vert^2_{Q^n} + \Vert c^2 \nabla u_1 \Vert^2_{Q^n} + \Vert \dt u_1 \Vert^2_{Q^n} \\
        &+ \Vert u_2 \Vert^2_{Q^n} + \int_{I_n} \sum_{K \in \mathcal{T}_h} h^2 \Vert c^2 u_1 \Vert^2_{H^2(K)} \dT \Big\}. 
    \end{aligned}
\end{equation}
%Note it suffices to show that $\tnorm{(\cdot,\cdot)}$ is a norm. We modify the steps from \cite{BP24} to show the following result.

@@

\begin{proof}
    Since $B[(\Uh,\Zh),(\Uh,-\Zh)] = \tnorm{(\Uh,\Zh)}^2 = \tnorm{(\Uh,\Zh)}\tnorm{(\Uh,-\Zh)}$, it directly follows that  
    \begin{align*}
        \sup_{(\Wh,\Yh) \in \ProdFullyDiscrSpace{k}{q} \times \ProdFullyDiscrSpace{k^\ast}{q^\ast}} &\frac{B[(\Uh,\Zh),(\Wh,\Yh)]}{\tnorm{(\Wh,\Yh)}} \\
        &\quad \ge \frac{B[(\Uh,\Zh),(\Uh,-\Zh)]}{\tnorm{(\Uh,-\Zh)}} = \tnorm{(\Uh,\Zh)}.
    \end{align*}
\end{proof}

@@

\begin{proof}
    Due to the assumption that $c_i \in C^\infty(\Omega_i)$, $i = 1,2$, we have that 
    \begin{equation*}
        \tnorm{(\Uh - \Vh, 0)}_{\wop} \le C \tnorm{(\Uh - \Vh, 0)}_{\square_1},
    \end{equation*}
    where $\tnorm{(\cdot, \cdot)}_{\square_1}$ is defined by \eqref{eq:tnormwop} with $c_i = 1$, $i = 1,2$. Thus, we can apply the interpolation results derived in \cite[Lem. 2.4]{BP24} to obtain the claim. 
\end{proof}

\noindent In the following, we want to show how the assumption of Lipschitz stability, cf. Assumption \ref{assum:LipschitzStability} or Thm. \ref{thm:Lipschitz} in the homogeneous case, can be used to show convergence in a physically relevant norm. To utilize these results, we need to deal with the fact that the discrete solution is discontinuous in time. To this end, let $L_{\Delta t} : \FullyDiscrSpace{k}{q} \rightarrow C^0(0,T;V_h^k)$ be a suitable lifting operator such that $L_{\Delta t} \ul_1 \in H^1(Q)$ as discussed in \cite[Sec. 4]{BP24}. Then, we formulate the following result.

>>>>

\affil[2]{\orgdiv{Institute for Numerical and Applied Mathematics}, \orgname{University of G\"{o}ttingen}, \orgaddress{\street{Lotzestr. 16-18}, \city{G\"{o}ttingen}, \postcode{37083}, \country{Germany}}}

@@

\section{Introduction}
\noindent In this article, we consider the discretization of a data assimilation problem for the wave equation in heterogeneous media. In the following, let $\Omega \subset \mathbb{R}^d$ be a bounded domain with smooth boundary $\partial \Omega$ such that there exists subdomains $\Omega_1, \Omega_2 \subseteq \Omega$ such that $\Omega = \overline{\Omega}_1 \cup \overline{\Omega}_2$ and $\Omega_1 \cap \Omega_2 = \emptyset$. We consider a scalar wave speed $c(x) = \mathbbm{1}_{\Omega_1} c_1(x) + \mathbbm{1}_{\Omega_2} c_2(x)$, $c_i \in C^\infty(\Omega_i)$, $i = 1,2$, which may be discontinuous across the interface $\Gamma = \Omega \setminus (\Omega_1 \cup \Omega_2)$, but constant in time. For a final time $T > 0$, we consider the space-time domain $Q = (0,T) \times \Omega$ with lateral boundary $\Sigma \coloneqq (0,T) \times \partial \Omega$. For any nonempty subset $\omega \subset \Omega$, we set $\omega_T \coloneqq (0,T) \times \omega$. In the following, let $\wop$ be the wave operator with discontinuous wave-speed defined as 
\begin{equation*}
    \wop u := \partial_t^2 - \div(c^2 \nabla u).
\end{equation*}
In particular, we write $\square_1$ if $c_i = 1$, $i = 1,2$, i.e. if we are considering the case of homogeneous media. 
Then, we consider the problem: find $u : Q \rightarrow \mathbb{R}$ such that 
\begin{equation}\label{eq:waveEquation}
        (\wop u, u) = (0,0) \quad \text{ in } Q \times \Sigma.  
\end{equation}
To ensure that the problem is well posed, one usually prescribes initial data
\begin{equation}\label{eq:initialData}
    (u,\dt u) \vert_{t = 0} = (u_0,u_1) \text{ in } \Omega, \tag{IVP} 
\end{equation}
where $u_0 \in H^1(Q)$ and $u_1 \in L^2(Q)$ are given. For a detailed analysis of the well-posedness of the problem \eqref{eq:waveEquation}-\eqref{eq:initialData} in heterogeneous media we refer to \cite{StolkPhD}. 
Here we consider a data assimilation problem instead, i.e. we assume that the initial data is unknown and that there are given measurements $u_{\omega} \in L^2(\omega_T)$ which are prescribed in a data domain $\omega_T$, $\omega \subset \Omega$, $\omega \not = \emptyset$:
\begin{equation}\label{eq:dataMatch}
    u = u_{\omega} \text{ in } \omega_T. \tag{DA}
\end{equation}

@@

\noindent For the heterogeneous case where $c_1 \not =  c_2$, the problem is less studied. In the absence of the GCC, we have to assume that the final time $T$ is large enough to ensure that problem \eqref{eq:waveEquation}-\eqref{eq:dataMatch} is well-posed. To be precise, Holmgren's unique continuation theorem guarantees that \eqref{eq:waveEquation}-\eqref{eq:dataMatch} is uniquely solvable if
\begin{equation}\label{eq:Tcondition}
    T > 2 \sup_{x \in \Omega} \operatorname{dist}_{c}(x,\omega_T),
\end{equation}
where $\operatorname{dist}_c(x,\omega_T) := \inf_{y \in \omega_T} d_c(x,y)$ with $d_c$ being the distance function defined as the length of the shortest continuous path in $\Omega$ connecting $x$ and $y$ with respect to the metric adapted to the discontinuous wave speed $c$. While the condition \eqref{eq:Tcondition} guarantees unique solvability, the stability of problem \eqref{eq:waveEquation}-\eqref{eq:dataMatch} can be poor. It has been shown \cite[Thm. 5.17]{Filippas22} under the assumption \eqref{eq:Tcondition} the following stability result holds:  
\begin{thm}[Thm. 5.17 of \cite{Filippas22}]\label{thm:filippas}
    Let $\omega_T \subset \Omega$ be nonempty and let $T$ be s.t. \eqref{eq:Tcondition} holds. Then, there exists $C,\kappa$ such that for any initial data $(u_0,u_1) \in H^1_0(\Omega) \times L^2(\Omega)$ and $u$ solving \eqref{eq:waveEquation}-\eqref{eq:initialData} one has for any $\mu > 0$ that 
    \begin{equation}\label{eq:FilippasEstimate}
        \Vert (u_0,u_1) \Vert_{L^2 \times H^{-1}} \le C e^{\kappa \mu} \left( \Vert u \Vert_{L^2(\omega_T)} + \Vert \wop u \Vert_{L^2(Q)} \right) + \frac{C}{\mu} \Vert (u_0,u_1) \Vert_{H^1 \times L^2}. 
    \end{equation}
    In particular, if $(u_0,u_1) \not = (0,0)$ we have that 
    \begin{equation}\label{eq:FilippasEstimateLog}
        \Vert (u_0,u_1) \Vert_{L^2 \times H^{-1}} \le C \frac{\Vert (u_0,u_1) \Vert_{H^1 \times L^2}}{\log \left( 1 + \frac{\Vert (u_0,u_1) \Vert_{H^1 \times L^2}}{ \Vert u \Vert_{L^2(\omega_T)} + \Vert \wop u \Vert_{L^2(Q)}} \right)}. 
    \end{equation} 
\end{thm}

@@

\section{Discretization (fully discrete \& dG in time)} 
\noindent In this section, we introduce a fully discrete discontinuous in time finite element method to discretize the data assimilation problem \eqref{eq:waveEquation}-\eqref{eq:dataMatch} for the wave equation in heterogeneous media. The discretization is based on the scheme introduced in \cite{BP24} for the homogeneous case. Before introducing the method in Section \ref{sec:method}, we describe the geometric partition of the space-time domain $Q$ into time slabs in Section \ref{sec:spaceTimeDiscretization}.

\subsection{Discretization of the space-time domain}\label{sec:spaceTimeDiscretization}
Let $\mathcal{T}_h$ be a quasi-uniform triangulation of $\Omega$ with mesh size $h$ such that both, the data domain $\omega$ and the interface $\Gamma$ are fitted. 
Note that the triangulation can be constructed such that for $h$ small enough the following continuous trace inequality holds \cite[Sec. 4.2]{BFMO21control}: For all $v \in [H^1(K)]^d$ and $K \in \mathcal{T}_h$, we have
\begin{equation}\label{eq:traceInequality}
    \Vert v \Vert_{[L^2(\partial K)]^d} \le C \left(h^{-1/2} \Vert v \Vert_{[L^2(K)]^d} + h^{1/2} \Vert \nabla v \Vert_{[L^2(K)]^d} \right).  
\end{equation}

@@

\section{Error analysis}
\noindent In this section, we analyze the approximation of the continuous problem \eqref{eq:waveEquation}-\eqref{eq:dataMatch} through \eqref{eq:discreteProblem}. In particular, we show that the error between the first component $\ul_1$ of the discrete solution and the continuous solution $u$ is bounded by the best approximation error in suitable discrete norms defined below. Then, we show that under Assumption \ref{assum:LipschitzStability}, we can recover convergence rates in a physically relevant norm.
Let $\vert \cdot \vert_{S_h}, \vert \cdot \vert_{\uparrow \downarrow}$, and $\Vert \cdot \Vert_{S_h^{\ast}}$ be the semi-norms, respectively norms, induced by the stabilizers $S_h$, $\Sud$, and $S_h^{\ast}$:
\begin{equation}
    \vert \Uh \vert^2_{S_h} := S_h(\Uh,\Uh), \quad \vert \Uh \vert^2_{\uparrow \downarrow} := \Sud(\Uh,\Uh), \quad \Vert \Zh \Vert^2_{S_h^{\ast}} := S_h^{\ast}(\Zh,\Zh).
\end{equation}
\noindent Then, we define the discrete norm 
\begin{equation}
    \tnorm{ (\Uh,\Zh) }^2 := \gamma \vert \Uh \vert^2_{S_h} + \vert \Uh \vert^2_{\uparrow \downarrow} + \Vert \ul_1 \Vert^2_{\omega_T} + \gamma^{\ast} \Vert \Zh \Vert^2_{S_h^\ast},
\end{equation}
and its strengthened version
\begin{equation}\label{eq:tnormwop}
    \begin{aligned}
        \tnorm{(\Uh, \Zh)}^2_{\wop} := \tnorm{(\Uh, \Zh)}^2 &+ \sum_{n = 0}^{N -1} \Big\{ \Vert \dt \ul_2 \Vert^2_{Q^n} + \Vert c^2 \nabla \ul_1 \Vert^2_{Q^n} + \Vert \dt \ul_1 \Vert^2_{Q^n} \\
        &+ \Vert \ul_2 \Vert^2_{Q^n} + \int_{I_n} \sum_{K \in \mathcal{T}_h} h^2 \Vert c^2 \ul_1 \Vert^2_{H^2(K)} \dT \Big\}. 
    \end{aligned}
\end{equation}
%Note it suffices to show that $\tnorm{(\cdot,\cdot)}$ is a norm. We modify the steps from \cite{BP24} to show the following result.

@@

\begin{proof}
    Since $B[(\Uh,\Zh),(\Uh,-\Zh)] = \tnorm{(\Uh,\Zh)}^2 = \tnorm{(\Uh,\Zh)}\tnorm{(\Uh,-\Zh)}$, we conclude that  
    \begin{align*}
        \sup_{(\Wh,\Yh) \in \ProdFullyDiscrSpace{k}{q} \times \ProdFullyDiscrSpace{k^\ast}{q^\ast}} &\frac{B[(\Uh,\Zh),(\Wh,\Yh)]}{\tnorm{(\Wh,\Yh)}} \\
        &\quad \ge \frac{B[(\Uh,\Zh),(\Uh,-\Zh)]}{\tnorm{(\Uh,-\Zh)}} = \tnorm{(\Uh,\Zh)}.
    \end{align*}
\end{proof}

@@

\begin{proof}
    Due to the assumption that $c_i \in C^\infty(\Omega_i)$, $i = 1,2$, there exists $C > 0$ s.t.  
    \begin{equation*}
        \tnorm{(\Uh - \Vh, 0)}_{\wop} \le C \tnorm{(\Uh - \Vh, 0)}_{\square_1},
    \end{equation*}
    where $\tnorm{(\cdot, \cdot)}_{\square_1}$ is defined by \eqref{eq:tnormwop} with $c_i = 1$, $i = 1,2$. Thus, we can apply the interpolation results derived in \cite[Lem. 2.4]{BP24} to obtain the claim. 
\end{proof}

\noindent In the following, we want to note that the assumption of Lipschitz stability, cf. Assumption \ref{assum:LipschitzStability} or Thm. \ref{thm:Lipschitz} in the homogeneous case, can be used to show convergence in a physically relevant norm. To utilize these results, we need to deal with the fact that the discrete solution is discontinuous in time. To this end, let $L_{\Delta t} : \FullyDiscrSpace{k}{q} \rightarrow C^0(0,T;V_h^k)$ be a suitable lifting operator such that $L_{\Delta t} \ul_1 \in H^1(Q)$ as introduced in \cite[Sec. 4]{BP24}. Then, we formulate the following result.
